---
title: "Multilateral"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multilateral}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=8
)
```





## Overview

The multilateral package provides one user facing function, that is `multilateral()`.
The user provides the necessary attributes of the dataset for the respective multilateral method.

This returns a list object containing the continuous spliced index, each individual windows calculation (useful for diagnostics), and splicing information if applicable.

```{r setup, echo=TRUE, error=FALSE, fig.keep='all', message=FALSE, warning=FALSE, results='hide'}
#devtools::install_github("MjStansfi/multilateral")
library(multilateral)

tpd_index <- multilateral(period = turvey$month,
                          id = turvey$commodity,
                          price = turvey$price,
                          quantity = turvey$quantity,
                          splice_method = "geomean",
                          window_length = 13,
                          index_method = "TPD")

```

```{r str of tpd_index}
str(tpd_index)
```
## Show case of index methods you can apply

The package provides two lazily loaded datasets, they are 'turvey' and 'synthetic_gfk'. The first dataset provides price and quantity information for products with a unique ID. The second dataset provides price and quantity information for products with a unique ID, however it also includes de-identified characteristics.

Note for this example we do not provide a window length, therefore the index is calculated over the whole dataset.

```{r turvey str}
str(turvey)
```

```{r turvey multilat, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)

index_methods <- c("TPD","GEKS-T","GEKS-J","GEKS-F")

start_run_time <- Sys.time()
turvey_multilats <- lapply(index_methods, function(index_method){
  
  
  temp_index <- multilateral(period = turvey$month,
                             id = turvey$commodity,
                             price = turvey$price,
                             quantity = turvey$quantity,
                             index_method = index_method)
  
  #For identification in plot
  temp_index$index$type <- index_method
  
  return(temp_index$index)  
})
end_run_time <- Sys.time()


turvey_multilats <- do.call(rbind,turvey_multilats)

plot <- ggplot(turvey_multilats)+geom_line(aes(x = period, y = index, colour = type))

print(plot)
```

Below indicates the time taken to run all four index methods above, highlighting efficieny of calculation
```{r run_time}
print(end_run_time-start_run_time)
```

### Data sources with features

```{r synthetic gfk}
str(synthetic_gfk)
```

Now we wrangle this into the necessary format, that is a unique observation for any given period.

```{r wrangle, error=FALSE, warning=FALSE, message=FALSE}
library(dplyr)

synthetic_gfk <- synthetic_gfk%>%
  group_by(month_num,prodid_num)%>%
  mutate(quantity = sum(quantity),
         value = sum(value))%>%
  ungroup()%>%
  unique

#Calculate the unit value (price)
synthetic_gfk$uv <- synthetic_gfk$value/synthetic_gfk$quantity

#Extract data.frame containing features of interest
features <-synthetic_gfk[,grepl("char",colnames(synthetic_gfk))]
```

Once the dataframe is in the correct format we can run it through the main function with index_method set to ‘GEKS-IT’.

Note GEKS-IT still requires a product id.

```{r run itrygeks, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}

itrygeks_index <- multilateral(period = synthetic_gfk$month_num,
                               id = synthetic_gfk$prodid_num, 
                               price = synthetic_gfk$uv,
                               quantity = synthetic_gfk$quantity,
                               features = features,
                               splice_method = "geomean",
                               index_method = "GEKS-IT")

#For identification in plot
itrygeks_index$index$type <- "itrygeks"
```
Alternatively we could run a basic TDH model, which does not require a unique ID.
```{r run TDH, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}

#Note no unique product ID

TDH_index <- multilateral(period = synthetic_gfk$month_num,
                              price = synthetic_gfk$uv,
                              quantity = synthetic_gfk$quantity,
                              features = features,
                              splice_method = "geomean",
                              index_method = "TDH")


#For identification in plot
TDH_index$index$type <- "TDH"
```

```{r plot features, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
feature_indexes <- rbind(itrygeks_index$index,TDH_index$index)

plot <- ggplot(feature_indexes)+geom_line(aes(x = period, y = index, colour = type))

print(plot)
```


## Show case of splice methods you can apply

Here we introduce a window length which therefore requires splicing

```{r turvey splice, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}

splice_methods <- c("geomean","geomean_short","window","movement","half")
turvey_splices <- lapply(splice_methods, function(splice_method){
  
  
  temp_index <- multilateral(period = turvey$month,
                             price = turvey$price,
                             id = turvey$commodity,
                             quantity = turvey$quantity,
                             window_length = 13,
                             splice_method = splice_method,
                             index_method = "TPD")
  
  temp_index$index$type <- splice_method
  
  return(temp_index$index)  
})

turvey_splices <- do.call(rbind,turvey_splices)

plot <- ggplot(turvey_splices)+geom_line(aes(x = period, y = index, colour = type))

print(plot)
```


## Why this package

* This package is fast relative to other available packages 
* This package allows flexibility of user input
* This package has the flexibility to incorporate new methods if needed
* The methods available are those that Stats NZ are either using or intending to explore the use of
* Building own package brings transparency and understanding to methods and nature of the functions

Below we compare the PriceIndices, IndexNumR, and multilateral package. We look at how you would calculate the same index, and how the speeds compare.

```{r results across package, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
library(multilateral)
library(IndexNumR)
library(PriceIndices)
library(dplyr)
#---------------------#multilateral package
turvey_multilateral <- multilateral(period = turvey$month,
                                    price = turvey$price,
                                    quantity = turvey$quantity,
                                    id = turvey$commodity,
                                    window_length = 13,
                                    splice_method = "geomean",
                                    index_method = "GEKS-F",
                                    check_inputs_ind = T)

#---------------------#PriceIndices package

#Requires data.frame with specific colnames, and format of time

turvey_mod <- turvey%>%select(time = month,
                              prodID = commodity,
                              prices = price,
                              quantities = quantity)

# turvey_priceindices <- geks_splice(turvey_mod, "1970-01", "1973-12", 13, splice = "mean", interval = T)


#---------------------#PriceIndices package

#Requires time to be a sequential numeric vector

turvey <- turvey%>%mutate(month_num = as.numeric(as.factor(month)))

turvey_IndexNumR <- GEKSIndex(turvey,
          pvar = "price",
          qvar = "quantity",
          pervar = "month_num",
          prodID = "commodity",
          indexMethod = "fisher",
          window = 13,
          splice = "mean")

```

```{r print resluts}
print(str(turvey_multilateral))
# print(str(turvey_priceindices))
print(str(turvey_IndexNumR))

```


```{r microbenchmark setup, include=FALSE}
library(multilateral)
library(IndexNumR)
library(PriceIndices)
library(dplyr)
#---------------------#multilateral package
turvey_geks_multilateral <- function(){multilateral(period = turvey$month,
                                    price = turvey$price,
                                    quantity = turvey$quantity,
                                    id = turvey$commodity,
                                    window_length = 13,
                                    splice_method = "geomean",
                                    index_method = "GEKS-F",
                                    check_inputs_ind = T)}

turvey_tpd_multilateral <- function(){multilateral(period = turvey$month,
                                    price = turvey$price,
                                    quantity = turvey$quantity,
                                    id = turvey$commodity,
                                    window_length = 13,
                                    splice_method = "geomean",
                                    index_method = "TPD",
                                    check_inputs_ind = T)}

#---------------------#PriceIndices package
turvey_geks_priceindices <- function(){geks_splice(turvey_mod, "1970-01", "1973-12", 13, splice = "mean", interval = T)}


turvey_tpd_priceindices <- function(){tpd_splice(turvey_mod, "1970-01", "1973-12", 13, splice = "mean", interval = T)}

#---------------------#PriceIndices package

turvey_geks_IndexNumR <- function(){GEKSIndex(turvey,
                                         pvar = "price",
                                         qvar = "quantity",
                                         pervar = "month_num",
                                         prodID = "commodity",
                                         indexMethod = "fisher",
                                         window = 13,
                                         splice = "mean")}

turvey_tpd_IndexNumR <- function(){WTPDIndex(turvey,
                                         pvar = "price",
                                         qvar = "quantity",
                                         pervar = "month_num",
                                         prodID = "commodity",
                                         window = 13,
                                         splice = "mean")}


```

```{r microbenchmark, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
library(microbenchmark)
#Converted to function, easier to see results
#Removed priceindices as too slow > 5min

speed_geks <- microbenchmark(turvey_geks_multilateral(),
                             # turvey_geks_priceindices(),
                             turvey_geks_IndexNumR(),
                             times = 1,
                            unit = 's')


speed_tpd <- microbenchmark(turvey_tpd_multilateral(),
                            # turvey_tpd_priceindices(),
                            turvey_tpd_IndexNumR(),
                            times = 1,
                            unit = 's')

```

```{r speed}
print(speed_geks)
print(speed_tpd)
```

Note the multilateral function also has a 'num_cores' parameter for parallelisation. Because calculation over a given window is not dependent on another, the majority of the function computation time - calculating indexes over each window - can be done in parallel. This is only applicable when there is more than one window, and only becomes noticebly quicker when dealing with larger datasets (>500k rows) as distributing to multiple cores has overheads. It is up to the user to find the correct balance.

```{r parallel, eval=FALSE, include=TRUE}
nrow(big_data)
#[1] 100000000

multilateral(period = big_data$month,
             price = big_data$price,
             quantity = big_data$quantity,
             id = big_data$commodity,
             window_length = 13,
             splice_method = "geomean",
             index_method = "TPD",
             check_inputs_ind = T,
             num_cores = 32)
```


